Eleven years of student replication projects provide evidence on the correlates of replicability in psychology

### 2023-04-05 paper/manuscript.Rmd
fix some citations and todos

A major concern in psychology, as in other fields, is that the published literature may not replicate as well as had previously been assumed. This is illustrated both by a few prominent findings that turn out to not replicate at all (CITE ex ego depletion, TMT), and by large-scale replication projects such as RP:P which found a replication rate around 40% for a sample of the psychology literature [@openscienceconsortium2015]. This low replication rates seems undesirable and seems to run counter to the idea of a cumulative science.

@@

Prior approaches to replicability have focused on interpreting results in terms of a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that "true" replicability may not be possible to estimate outside of a specific sample (CITE van bavel etc.) or even time period (CITE ramscar).

Further, the methods for estimating this quantity have been theoretically problematic. Sampling schemes for prior work typically do not reflect an entirely random sample from the literature; instead they sample from specific journals where results may be of more interest and adjust the sample for feasibility concerns. These are reasonable sampling choices, but they undermine the claim that the estimand is the level of "truth" in the literature as a whole. Sampling truely at random from the literature may not even be desirable, as  arguably a literature will succeed if useful discoveries come out of it, not if random findings are true (CITE "Science is not signal detection" ref). How much we care about whether a study is "trustworthy" is not uniformly distributed across the literature.

@@

## Single predictors
Properties of both the original study and the replication can influence whether or not the replication is a success. We chose a set of predictor variables from the correlational results of RP:P and our own intuitions about experimental factors that might impact replication success as well as some covariates related to how close the replication would be. A full description of these features is given in methods.

Many of these predictors individually correlate with subjective replication success (Table \ref{tab:cor}). Predictors of higher replicability included within-subjects designs, higher numbers of trials, and open data. Predictors of lower replicability included Single vignetted studies, social psychology studies, and original-replication pairs where the replication switched to online.

Distributions of study outcomes across some of these properties are shown in Figure \ref{fig:predictors-graph}. Both social and cognitive psychology studies were well represented, and the cognitive psychology studies replicated at twice the rate of social psychology studies. Within and between subjects designs were both common, and within replicated four times as much. Similarly, studies with multiple vignettes replicated 1.5 times more than single vignetted studies. However, there were strong correlations among these experimental features and between these experimental features and subfield. TODO actual numbers here!

@@

Nearly all replications studies were conducted online, but original studies were split between using in-person and online recruitment. Replications that switched to online were less likely to replicate than those that had the same modality as the original (generally both online, in a few cases both in-person).  While online studies in general show comparable results to studies conducted in person TODO CITATIONS, switching the modality does decrease the closeness of the replication, and some studies done in person may not have been well adapted (ex. inductions may be weaker or attention checks inadequate to the new sample).

@@

## Regression model
While a number of the predictors show individual correlations with the subjective replication score, many of the predictors are also correlated with one another. In order to determine which predictors were the strongest, we ran a pre-registered regularized regression model (see Methods for details). The coefficient estimates are shown in Figure \ref{fig:mod-results}. Due to a large number of predictors coupled with a small and noisy dataset, even with strong regularization, there is much uncertainty around the coefficients. The general directions of coefficients are consistent with the effects of the predictors in isolation. Within-subjects designs seem an especially strong indicator of replicability TODO include estimate and range in text!

@@

Other replication projects have other estimands: @camerer2018 seems to ask something closer to how likely studies are to replicate when preformed by an expert with a large budget, and @ML5 TODOCITE asks how likely certain studies are to replicate when performed with extensive feedback from the original authors.

@@

Our results are limited by the number and quality of the studies we included. These studies are not necessarily representative of the literature as a whole. TODO what are the limitations we want?

@@

## Dataset
The dataset of replication projects comes from class projects conducted in PSYCH 251 (earlier called PSYCH 254) a graduate-level experimental methods class taught by MCF from 2011 to 2022. This class is commonly taken by first year graduate students in psychology and related disciplines, and it has been a requirement of the Psychology PhD since around 2015. Each student chose a study to replicate, implemented the study, wrote analysis code, pre-registered their replication, ran the study, and turned in a structured final report including methods, analytic plan, changes from the original study, confirmatory and exploratory analyses, and discussion of outcomes. Students were encouraged to do experimental replications, but some students chose to replicate correlational outcomes or do computational reproducibility projects instead. We cannot include the full student reports for confidentiality reasons, but we include an example as well as the template given to students in the repo. TODO example and template

Students were free to choose what study they replicated; the recommended path for students who did not have their own ideas was to pick an interesting study from a recent year of Psychological Science (this led to a high fraction of Psych Science articles in the replication sample TODO FOOBAR %).

We note that 4 (TODO check) of the replication projects were included in RP:P, and FOOBAR of them were previously reported in @hawkins.

>>>>

A major concern in psychology, as in other fields, is that the published literature may not replicate as well as had previously been assumed. This is illustrated both by a few prominent findings that turn out to not replicate at all [ex. terror management theory; @klein2022], and by large-scale replication projects such as RP:P which found a replication rate around 40% for a sample of the psychology literature [@openscienceconsortium2015]. This low replication rates seems undesirable and seems to run counter to the idea of a cumulative science.

@@

Prior approaches to replicability have focused on interpreting results in terms of a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that "true" replicability may not be possible to estimate outside of a specific sample [@vanbavel2016] or even time period [@ramscar].

Further, the methods for estimating this quantity have been theoretically problematic. Sampling schemes for prior work typically do not reflect an entirely random sample from the literature; instead they sample from specific journals where results may be of more interest and adjust the sample for feasibility concerns. These are reasonable sampling choices, but they undermine the claim that the estimand is the level of "truth" in the literature as a whole. Sampling truely at random from the literature may not even be desirable, as  arguably a literature will succeed if useful discoveries come out of it, not if random findings are true [@wilson2020]. How much we care about whether a study is "trustworthy" is not uniformly distributed across the literature.

@@

```{r}

 within <- glm(sub_rep ~ is_within, data=for_cor, family=binomial)$coefficient[2] |> exp() 
 vignette <- glm(sub_rep ~ single_vignette, data=for_cor, family=binomial)$coefficient[2] |> exp()

  subfield <- glm(sub_rep ~ subfield, data=for_cor |> filter(subfield %in% c("social", "cognitive")), family=binomial)$coefficient[2] |> exp()

```
## Single predictors
Properties of both the original study and the replication can influence whether or not the replication is a success. We chose a set of predictor variables from the correlational results of RP:P and our own intuitions about experimental factors that might impact replication success as well as some covariates related to how close the replication would be. A full description of these features is given in methods.

Many of these predictors individually correlate with subjective replication success (Table \ref{tab:cor}). Predictors of higher replicability included within-subjects designs, higher numbers of trials, and open data. Predictors of lower replicability included single vignetted studies, social psychology studies, and original-replication pairs where the replication switched to online.

Distributions of study outcomes across some of these properties are shown in Figure \ref{fig:predictors-graph}. Both social and cognitive psychology studies were well represented, and the cognitive psychology studies replicated at `r 1/subfield |> round(2)` times the rate of social psychology studies. Within and between subjects designs were both common, and within replicated `r within |> round(2)` times as much. Similarly, studies with multiple vignettes replicated `r 1/vignette |> round(2)` times more than single vignetted studies. However, there were strong correlations among these experimental features and between these experimental features and subfield.

@@

Nearly all replications studies were conducted online, but original studies were split between using in-person and online recruitment. Replications that switched to online were less likely to replicate than those that had the same modality as the original (generally both online, in a few cases both in-person).  While online studies in general show comparable results to studies conducted in person [@crump2013]TODO MORE CITATIONS, switching the modality does decrease the closeness of the replication, and some studies done in person may not have been well adapted (ex. inductions may be weaker or attention checks inadequate to the new sample).

@@

```{r}
tier1_subjective <- read_rds(here(model_location, "tier1_subjective.rds"))

summary(tier1_subjective)$fixed |> as_tibble(rownames="var") |> filter(var=="is_within")|> mutate(text=str_c(round(Est.Error,2), "[", round(`l-95% CI`,2),",", round(`u-95% CI`,2),"]")) |> pull(text) ->within_range
```

## Regression model
While a number of the predictors show individual correlations with the subjective replication score, many of the predictors are also correlated with one another. In order to determine which predictors were the strongest, we ran a pre-registered regularized regression model (see Methods for details). The coefficient estimates are shown in Figure \ref{fig:mod-results}. Due to a large number of predictors coupled with a small and noisy dataset, even with strong regularization, there is much uncertainty around the coefficients. The general directions of coefficients are consistent with the effects of the predictors in isolation. Within-subjects designs seem an especially strong indicator of replicability (`r within_range`).

@@

Other replication projects have other estimands: @camerer2018 seems to ask something closer to how likely studies are to replicate when preformed by an expert with a large budget, and @ebersole2020 asks how likely certain studies are to replicate when performed with extensive feedback from the original authors.

@@

Our results are limited by the number and quality of the studies we included. These studies are not necessarily representative of the studies of interest to psychologists as a whole. The replication studies were not designed from the start with this analysis in mind, so analysis is limited by the choices and reporting used at the time of replications. TODO what are the limitations we want?

@@

## Dataset

```{r}
all <- read_csv(here("data", "raw_data.csv")) |> filter(include %in% c("stats", "pred_int", "exp"))

ps <- all|> mutate(journal=str_to_lower(journal)) |>filter(journal=="psych sci")

```
The dataset of replication projects comes from class projects conducted in PSYCH 251 (earlier called PSYCH 254) a graduate-level experimental methods class taught by MCF from 2011 to 2022. This class is commonly taken by first year graduate students in psychology and related disciplines, and it has been a requirement of the Psychology PhD since around 2015. Each student chose a study to replicate, implemented the study, wrote analysis code, pre-registered their replication, ran the study, and turned in a structured final report including methods, analytic plan, changes from the original study, confirmatory and exploratory analyses, and discussion of outcomes. Students were encouraged to do experimental replications, but some students chose to replicate correlational outcomes or do computational reproducibility projects instead. We cannot include the full student reports for confidentiality reasons, but we include an example as well as the template given to students in the repo. TODO example and template

Students were free to choose what study they replicated; the recommended path for students who did not have their own ideas was to pick an interesting study from a recent year of Psychological Science (this led to a high fraction of Psych Science articles in the replication sample, `r nrow(ps)`, `r nrow(ps)/nrow(all)*100 |> round()`% of studies).

We note that 4 (TODO check) of the replication projects were included in RP:P, and 10 of them were previously reported in @hawkins.

### 2023-04-05 paper/manuscript.tex
fix some citations and todos

A major concern in psychology, as in other fields, is that the published literature may not replicate as well as had previously been assumed. This is illustrated both by a few prominent findings that turn out to not replicate at all (CITE ex ego depletion, TMT), and by large-scale replication projects such as RP:P which found a replication rate around 40\% for a sample of the psychology literature (\protect\hyperlink{ref-openscienceconsortium2015}{Consortium 2015}). This low replication rates seems undesirable and seems to run counter to the idea of a cumulative science.

@@

Prior approaches to replicability have focused on interpreting results in terms of a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that ``true'' replicability may not be possible to estimate outside of a specific sample (CITE van bavel etc.) or even time period (CITE ramscar).

Further, the methods for estimating this quantity have been theoretically problematic. Sampling schemes for prior work typically do not reflect an entirely random sample from the literature; instead they sample from specific journals where results may be of more interest and adjust the sample for feasibility concerns. These are reasonable sampling choices, but they undermine the claim that the estimand is the level of ``truth'' in the literature as a whole. Sampling truely at random from the literature may not even be desirable, as arguably a literature will succeed if useful discoveries come out of it, not if random findings are true (CITE ``Science is not signal detection'' ref). How much we care about whether a study is ``trustworthy'' is not uniformly distributed across the literature.

@@

Many of these predictors individually correlate with subjective replication success (Table \ref{tab:cor}). Predictors of higher replicability included within-subjects designs, higher numbers of trials, and open data. Predictors of lower replicability included Single vignetted studies, social psychology studies, and original-replication pairs where the replication switched to online.

Distributions of study outcomes across some of these properties are shown in Figure \ref{fig:predictors-graph}. Both social and cognitive psychology studies were well represented, and the cognitive psychology studies replicated at twice the rate of social psychology studies. Within and between subjects designs were both common, and within replicated four times as much. Similarly, studies with multiple vignettes replicated 1.5 times more than single vignetted studies. However, there were strong correlations among these experimental features and between these experimental features and subfield. TODO actual numbers here!

@@

Nearly all replications studies were conducted online, but original studies were split between using in-person and online recruitment. Replications that switched to online were less likely to replicate than those that had the same modality as the original (generally both online, in a few cases both in-person). While online studies in general show comparable results to studies conducted in person TODO CITATIONS, switching the modality does decrease the closeness of the replication, and some studies done in person may not have been well adapted (ex. inductions may be weaker or attention checks inadequate to the new sample).

@@

While a number of the predictors show individual correlations with the subjective replication score, many of the predictors are also correlated with one another. In order to determine which predictors were the strongest, we ran a pre-registered regularized regression model (see Methods for details). The coefficient estimates are shown in Figure \ref{fig:mod-results}. Due to a large number of predictors coupled with a small and noisy dataset, even with strong regularization, there is much uncertainty around the coefficients. The general directions of coefficients are consistent with the effects of the predictors in isolation. Within-subjects designs seem an especially strong indicator of replicability TODO include estimate and range in text!

@@

Other replication projects have other estimands: Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) seems to ask something closer to how likely studies are to replicate when preformed by an expert with a large budget, and (\protect\hyperlink{ref-ML5}{\textbf{ML5?}}) TODOCITE asks how likely certain studies are to replicate when performed with extensive feedback from the original authors.

@@

Students were free to choose what study they replicated; the recommended path for students who did not have their own ideas was to pick an interesting study from a recent year of Psychological Science (this led to a high fraction of Psych Science articles in the replication sample TODO FOOBAR \%).

We note that 4 (TODO check) of the replication projects were included in RP:P, and FOOBAR of them were previously reported in Hawkins et al. (\protect\hyperlink{ref-hawkins}{n.d.}).

@@

@@

@@

@@

@@

>>>>

A major concern in psychology, as in other fields, is that the published literature may not replicate as well as had previously been assumed. This is illustrated both by a few prominent findings that turn out to not replicate at all {[}ex. terror management theory; Klein et al. (\protect\hyperlink{ref-klein2022}{2022}){]}, and by large-scale replication projects such as RP:P which found a replication rate around 40\% for a sample of the psychology literature (\protect\hyperlink{ref-openscienceconsortium2015}{Consortium 2015}). This low replication rates seems undesirable and seems to run counter to the idea of a cumulative science.

@@

Prior approaches to replicability have focused on interpreting results in terms of a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that ``true'' replicability may not be possible to estimate outside of a specific sample (\protect\hyperlink{ref-vanbavel2016}{Van Bavel et al. 2016}) or even time period (\protect\hyperlink{ref-ramscar}{Ramscar et al. n.d.}).

Further, the methods for estimating this quantity have been theoretically problematic. Sampling schemes for prior work typically do not reflect an entirely random sample from the literature; instead they sample from specific journals where results may be of more interest and adjust the sample for feasibility concerns. These are reasonable sampling choices, but they undermine the claim that the estimand is the level of ``truth'' in the literature as a whole. Sampling truely at random from the literature may not even be desirable, as arguably a literature will succeed if useful discoveries come out of it, not if random findings are true (\protect\hyperlink{ref-wilson2020}{Wilson et al. 2020}). How much we care about whether a study is ``trustworthy'' is not uniformly distributed across the literature.

@@

Many of these predictors individually correlate with subjective replication success (Table \ref{tab:cor}). Predictors of higher replicability included within-subjects designs, higher numbers of trials, and open data. Predictors of lower replicability included single vignetted studies, social psychology studies, and original-replication pairs where the replication switched to online.

Distributions of study outcomes across some of these properties are shown in Figure \ref{fig:predictors-graph}. Both social and cognitive psychology studies were well represented, and the cognitive psychology studies replicated at 2.4390244 times the rate of social psychology studies. Within and between subjects designs were both common, and within replicated 3.35 times as much. Similarly, studies with multiple vignettes replicated 2.6315789 times more than single vignetted studies. However, there were strong correlations among these experimental features and between these experimental features and subfield.

@@

Nearly all replications studies were conducted online, but original studies were split between using in-person and online recruitment. Replications that switched to online were less likely to replicate than those that had the same modality as the original (generally both online, in a few cases both in-person). While online studies in general show comparable results to studies conducted in person (\protect\hyperlink{ref-crump2013}{Crump et al. 2013})TODO MORE CITATIONS, switching the modality does decrease the closeness of the replication, and some studies done in person may not have been well adapted (ex. inductions may be weaker or attention checks inadequate to the new sample).

@@

While a number of the predictors show individual correlations with the subjective replication score, many of the predictors are also correlated with one another. In order to determine which predictors were the strongest, we ran a pre-registered regularized regression model (see Methods for details). The coefficient estimates are shown in Figure \ref{fig:mod-results}. Due to a large number of predictors coupled with a small and noisy dataset, even with strong regularization, there is much uncertainty around the coefficients. The general directions of coefficients are consistent with the effects of the predictors in isolation. Within-subjects designs seem an especially strong indicator of replicability (0.55{[}-0.01,2{]}).

@@

Other replication projects have other estimands: Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) seems to ask something closer to how likely studies are to replicate when preformed by an expert with a large budget, and Ebersole et al. (\protect\hyperlink{ref-ebersole2020}{2020}) asks how likely certain studies are to replicate when performed with extensive feedback from the original authors.

@@

Students were free to choose what study they replicated; the recommended path for students who did not have their own ideas was to pick an interesting study from a recent year of Psychological Science (this led to a high fraction of Psych Science articles in the replication sample, 80, 45.4545455\% of studies).

We note that 4 (TODO check) of the replication projects were included in RP:P, and 10 of them were previously reported in Hawkins et al. (\protect\hyperlink{ref-hawkins}{n.d.}).

@@

\leavevmode\vadjust pre{\hypertarget{ref-crump2013}{}}%
Crump MJC, McDonnell JV, Gureckis TM (2013) Evaluating {Amazon}'s {Mechanical Turk} as a {Tool} for {Experimental Behavioral Research} (S Gilbert, Ed.). \emph{PLoS ONE} \textbf{8}:e57410. doi:\href{https://doi.org/10.1371/journal.pone.0057410}{10.1371/journal.pone.0057410}

@@

\leavevmode\vadjust pre{\hypertarget{ref-klein2022}{}}%
Klein RA, Cook CL, Ebersole CR, Vitiello C, Nosek BA, Hilgard J, Ahn PH, Brady AJ, Chartier CR, Christopherson CD, Clay S, Collisson B, Crawford JT, Cromar R, Gardiner G, Gosnell CL, Grahe J, Hall C, Howard I, Joy-Gaba JA, Kolb M, Legg AM, Levitan CA, Mancini AD, Manfredi D, Miller J, Nave G, Redford L, Schlitz I, Schmidt K, Skorinko JLM, Storage D, Swanson T, Van Swol LM, Vaughn LA, Vidamuerte D, Wiggins B, Ratliff KA (2022) Many {Labs} 4: {Failure} to {Replicate Mortality Salience Effect With} and {Without Original Author Involvement}. \emph{Collabra: Psychology} \textbf{8}:35271. doi:\href{https://doi.org/10.1525/collabra.35271}{10.1525/collabra.35271}

@@

\leavevmode\vadjust pre{\hypertarget{ref-ramscar}{}}%
Ramscar M, Shaoul C, Baayen RH Why many priming results don't (and won't) replicate: {A} quantitative analysis.

@@

\leavevmode\vadjust pre{\hypertarget{ref-vanbavel2016}{}}%
Van Bavel JJ, Mende-Siedlecki P, Brady WJ, Reinero DA (2016) Contextual sensitivity in scientific reproducibility. \emph{Proc Natl Acad Sci} \textbf{113}:6454--6459. doi:\href{https://doi.org/10.1073/pnas.1521897113}{10.1073/pnas.1521897113}

@@

\leavevmode\vadjust pre{\hypertarget{ref-wilson2020}{}}%
Wilson BM, Harris CR, Wixted JT (2020) Science is not a signal detection problem. \emph{Proc Natl Acad Sci USA} \textbf{117}:5559--5567. doi:\href{https://doi.org/10.1073/pnas.1914237117}{10.1073/pnas.1914237117}
